Constraining the High Density Nuclear Symmetry Energy with Pions

### 2020-06-26 chapters/ch_data_analysis.tex
Bill edits added Experiment section
Halfway through data analysis 1 backup

The S$\pi$RITROOT software is modular code based on the FAIRROOT package written in C++ \cite{fairroot}. The code is separated into modules which can be turned on or off depending on the users needs. The base classes and frame work is provided by the FAIRROOT package, while the specifics of each detector need to be added into the base classes. In the following an outline of the main tasks in the S$\pi$RITROOT software will be briefly discussed with the details of each task discussed later. The discussion of the software is really a practical, base understanding, appropriate for this thesis. For more details about the tracking algorithm the reader is refered to \cite{spiritroot_paper}, and to actual software which is published on GitHub \cite{spiritroot_git}.

@@

\subsection{Decoder task}
The decoder task reads in the raw data output from the GET electronics and separates the data from the headers using the GETDecoder \cite{getdecoder}. A map is stored mapping each channel in the CoBo board onto the physical pad plane, the information is stored in a pad container called the STPad class. This container stores the row, layer, and the raw ADC time bucket spectrum. All of the pads combined form the event container, STRawEvent class, which represents the raw event information.

Three basic calibrations are performed at the raw data level within this task. The gain calibration calibrates the electronics will be described in Section\ref{sec:elecCalib} and the anode gain calibration is discussed in Section~\ref{sec:anodeCalib}. The gating grid noise subtraction and time bucket cut which is described in Section~\ref{sec:ggnoisesub}.
%gain calibration 
%ggnoise subtraction
%GG time bucket cut

\subsection{Pulse Shape Algorithm (PSA) Task}
\label{sec:psa}
%Yoffset match the TPC-Vertex_y with BDC_Y
%YPedestalOffset what is it for?
%cobo gain matching
An algorithm loops over the raw time bucket spectrum to find all the unique pules in each pad. It is very likely that there are several pulses in a pad coming from multiple tracks passing under the same pad which are separated only in arrival time. Figure~\ref{fig:psaTask} shows the PSA task working on one particular pad in experimental data. The shaded histogram in every sub-figure represents the ADC raw spectrum of the pad. In Fig.~\ref{fig:psa0} we see the PSA has identified the first pulse and fitted it with the standard pulse, the result of the fit is shown by the red line. The fitted red line is then subtracted from the ADC spectrum, which is shown in the shaded histogram in Fig.~\ref{fig:psa1}. Here, the second pulse is found and fitted again with the same procedure, the red line will be subtracted from the raw ADC spectrum. The black line represents the sum of all the fitted pules in each step. This continues until all pules are found or until the PSA reaches the last bin in the spectrum. Figure~\ref{fig:psa4} shows the final result where 4 fitted pulses were found. The agreement between the total summed line and the raw ADC spectrum exemplifies how successful the PSA is in this particular set of data.

@@

The fit of each individual pulse contains the height of the pulse and the timing information of the start of the pulse, defined to be the location in time of 10\% of the rising edge. The height of the pulse, which is related to the charge induced on the pad, $Q$, and the timing information $t_o$, is saved into a new container representing a ``hit", called the STHit class. After the PSA processes each pad in the raw event, an entire hit cloud of the event is produced. The x and z-coordinates come from the center of the pad which is already stored in the map of STPad. Since the STHit class inherits from the STPad class this information is also carried over. The y-coordinate is calculated using the known drift velocity $v$ and time of the signal $t_o$ as $y = v\cdot t_0$, $t_0$. The hit class forms the fundamental structure from which all tracking is performed since it contains the x,y,z spacial coordinates, and charge information $Q$.

@@

 The function of the  helix track finding algorithm is to sort hit cloud into unique sets of hits representing tracks. This is performed by a Riemann track finding algorithm where the Cartesian space is mapped onto a Riemann sphere. This is a particular transformation in which unique helices in the Cartesian space form great circles on the Riemann sphere. This is particularly useful since hits which correspond to a track form a helices the Cartesian space of the TPC coordinates. It is difficult to search for tracks in the hit cloud by fitting a general helix to different combinations of hits. It is much easier to search for collection of hits by fitting a plane to the great circles which are formed on the Riemann sphere. From here the set of hits which form a unique track are stored into a new helix track container called the STHelixTrack class.

\paragraph{Definition of clustering}
The hits within a helix track are dense and the position resolution of an individual hit is not very precise. As mentioned in Section~\ref{sec:prf}, the TPC can achieve sub-millimeter position resolution by utilizing the PRF. In the helix track finding task, the hits are further reduced into more precise clusters. A brief description of the method of clustering is illustrated in Figure \ref{fig:topview}. As will be seen, it is sufficient to simply cluster hits along one direction, either along the x or z-direction, which ever is most perpendicular to the local track region. For example, the three clusters at the bottom of Figure \ref{fig:topview} are clustered along the x-axis, and the upper three are along the z-axis. The bold pads highlight the hits belonging to an example cluster from each set. Since we decide to cluster only in one direction, there is a natural inflection point in a track at which the direction of clustering must switch. This depends on the crossing angle of the track $\theta$, which is defined as the angle between the tangent of the track  at that cluster and the x-axis. In this definition $\theta = \ang{90}$ when the track is going along the z-axis and $\theta = \ang{0}$ for a track going along the x-axis.  The clustering direction in the case $\ang{45} < \theta \leq \ang{90} $ is along the x-direction. For $\ang{0} < \theta \leq \ang{45}$ we cluster along the z-axis.

@@

 The position along the clustering direction, $\overline{X}$, is calculated as the charge-weighted average, $\overline{X} = \sum_i q_ix_i$, where $q_i$ and $x_i$ are the charge and position of the i-th hit in the cluster. The other direction is set to the center of the pad. For example, if we are clustering hits along the x-axis, the z-position is set to the center of the pad in the z-direction and vice versa. Switching the clustering direction gives better position resolution than if we clustered only along one direction. You can imagine if we calculated the clusters only along the x-axis for tracks with $\theta \approx 0^{\circ}$ the x-position is not well defined. The cluster position and charge is stored into a new container represented the cluster called the STHitCluster class. All the clusters which belong to a particular helix track is also stored in the corresponding STHelixTrack class.

\subsection{Correction Task}
%extend dynamic range
%PRF cut
%space charge correciton
In this task the hit distribution, within a given cluster, is checked against the measured PRF discussed in Section~\ref{sec:prf}. Figure~\ref{fig:prfpimData} shows black lines around the measured PRF in the data, which form gates defining the acceptable PRF region. If less than 50\% of the hits within a cluster lie within this gate, the cluster is thrown out from the track. A majority of the clusters contains either 2 or 3 hits in total; in the case of 2 total hits, at least one of them needs to lie in the gate, in the case of 3 total hits, at least 2 hits need to lie within the gate. Clusters which do not follow this PRF gate usually have been corrupted. Typically charges from other nearby tracks is the typical way clusters become corrupt.

@@

This correction task also handles several other important corrections. The first is a novel algorithm which extends the dynamic range of the electronics and is discussed in detail in Section~\ref{sec:extendDynamicRange}. The other correction task that is handled here is the space charge correction which will be discussed in more detail in Section~\ref{sec:spacecharge}.

\subsection{Momentum and track reconstruction (GENFIT)}
After all of the corrections are applied to the clusters, the cluster's position and charge values are passed to a software package which reconstructs the momentum of the track, called GENFIT \cite{genfit}. The details of Kalman fitting algorithm used in the GENFIT reconstruction are left for the reader here \cite{genfit}. GENFIT returns a momentum and charge value for the track which is stored into a new track container called the STRecoTrack class.

The task is repeated a second time, storing a second copy of the track class but this time with the BDC vertex point added as a constraining cluster. The momentum resolution is greatly improved by adding the vertex as a constraining point; since the BDC detectors, described in Section~\ref{sec:bdc}, give such an accurate vertex location. The energy loss of the track is calculated by the truncated mean method described in Eq.~\ref{eq:truncmean}. This is the final part of the analysis which pertains to tracks, and the PID can be constructed using the energy loss, momentum, and charge information.

\subsection{Vertex tracking (RAVE)}
\label{sec:vertex}
After all tracks are reconstructed by GENFIT, the tracks are then passed to the RAVE software package which reconstructs the event vertex from a collection of tracks \cite{rave}. Only tracks which are THIS CONDITION are used in making the event vertex. The vertex location along the beam axis determines whether or not the beam collided with the target.

@@

\subsection{Gating grid noise subtraction}
\label{sec:ggnoisesub}
The opening of the gating grid is essentially shorting adjacent wires in the grid together, allowing them to reach equilibrium as fast as possible. Since the impedance of both sides was not entirely matched properly this caused an imbalance of charge which bounce back and forth in the grid in an under-dampened manner. This oscillating charged induced  residual induced signals early in the time bucket spectrum creating extra source of noise in the data. Figure~\ref{fig:ggNoiseSub} shows in the upper panel the gating grid noise ADC time bucket spectrum for 2000 events in a given pad. The gating grid noise is stable for a given pad and the mean value --shown as the red line-- can be calculated after averaging over several thousands events. The raw gating grid noise lasts for 100 Tbs extending into the real data with a exponentially decreasing ampli

 %The ground grid shielded some of the noise, but the path to ground was on two ends and was not sufficiency for this frequency.

@@

The bottom panel of Fig.~\ref{fig:ggNoiseSub} shows the gating grid noise self correcting itself; a best case scenario. We can see the pedestal, which is around 3800 ADC in the upper pannel,  is automatically taken into account when subtracting the mean profile, and the subtracted ADC spectra is centered around 0. The residual gating grid noise that was not canceled by the subtraction remains, but is significantly reduced to a much smaller time bucket region. Most of the residual noise happens in regions where the derivative of the gating grid noise is large. In these areas the ADC value is dependent upon small jitters in the timing of the signal and as a result varies wildly. The residual noise in the subtracted spectra is removed by cutting away all the bins below 30 tb.

In the experiment the gating grid itself was quite stable in the  ${}^{132}$Sn and ${}^{124}$Sn systems and was not very stable in the ${}^{108}$Sn and ${}^{112}$Sn beams, where the gating grid broke several times. The gating grid noise was monitored and we regularly took gating grid noise profiles, especially after a gating grid board was replaced.

@@

A cocktail of several light charged particles was produced (p,d,t,${}^{3}$He,${}^{4}$He), and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BigRIPS spectrometer was set so that the momentum resolution of the beam was $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied, with an empty target. A thick Aluminum target was also inserted to provide a slightly lower energy point for part of the lower rigidity setting, effectively creating three energy calibration points over several particle species. In some cases the certain particles were not able to propagate or were strongly attenuated.

Since the cocktail beam was a mono-energetic source of particles and the actual momentum resolution was less than 1\%, the observed momentum resolution can be interpreted as the intrinsic TPC momentum resolution. The momentum resolution in general depends on several factors such as the particle's angle, momentum, charge, track multiplicity, etc. This calibration beam represents and ideal situation where the track was parallel to the pad plane and only one particle was measured at one time. The average momentum resolution was measured to be $dp/p = 2\%$ over the full range of particle species for all settings.

The energy loss resolution can also be directly measured since each energy setting represents a monochromatic source of each particle species, which have a well defined energy loss distribution. The energy loss resolution was measured to be 5\% for all particles over the energy setting measured and is summarized in Table~\ref{tb:momresolution}.

Since the magnetic dipole setting of the BigRIPS spectrometer is well defined, we can calculate the expected momenta of each particle species measured. The enegy losses through various materials in the beam line were propagated using LISE++ software \cite{lise++}. The measured momenta of the calibration beam differed significantly from the expected values -- up to 5\% in the high momentum calibration -- as seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. This effect is attributed to inhomogenatities in the magnetic field which introduces electron drift velocities in the direction of $\vec{E}\times\vec{B}$ direction as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality.  The details of the correcting for the $\vec{E}\times\vec{B}$ effect will be discussed later in Section~\ref{sec:spacecharge} in a  general way which also includes the correction for the positive space charge. The same correction technique was applied here where the cocktail beam is the special case of zero space charge, only having $\vec{E}\times\vec{B}$ components.

The values under the corrected column in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}, represent the corrected data. A significant improvement is seen in the high momentum setting going from around ~5\% disagreement to within ~1\% agreement in the corrected data. For the lower momentum settings (Tables~\cref{tb:cocktail100tar,tb:cocktail100}), protons see a slight improvement of about ~1\% where as the deuterons are over corrected in both settings. The level of agreement of the all corrected values is still well within the estimated momentum resolution of the TPC.

@@

The electronics were calibrated by measuring the response of each channel to an input signal supplied by a pulse generator. This is a relative calibration technique with the intent to calibrate the varying gains in each channel relative to one another, and not an absolute calibration. The pulse was distributed to all the electronics channels by pulsing the ground plane for a range of input voltages. This distributed the pulse evenly across the entire pad plane. The input voltage is plotted as a function of the measured height of the pulse in the electronics channel given in units of ADC in Fig.~\ref{fig:gaincalib}, for every channel. The small variation in each channel can be seen as the wide band around each measurement point. A linear fit is performed for each channel to get the calibration line for each channel. A channel is chosen which provides a reference calibration to which all other channels are calibrated too. This is done by matching the slopes of each channel to the reference calibration slope.  The right panel shows the resulting distribution of channels after calibration.

\subsection{Anode gain calibration}
\label{sec:anodeCalib}
As discussed in Section~\ref{sec:wireplanes} the voltage of the anode sections 12 and 14 were reduced during the experiment due to high currents being observed on the wires. Out of all the runs used in the analysis in this thesis \ref{tb:runList}, the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all the other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, we increase the gain of the pads which lie above these anode wires  in the software gain of the other channels. The gain factor is found by plotting the energy loss in the high gain sections to the low gain sections.

@@

 A map of the channels which are affected by the low gain and the relevant gain factor was input into the software. In the decoder task the raw ADC spectra was multiplied by the input gain for a given channel. The thresholds of those channels also were multiplied by the same gain factor since the noise levels were also artificially magnified by this method.

@@

Using a TPC for measurements of HIC in nuclear physics presents a different set of challenges as opposed to higher energy experiments. Typically in higher energy experiments fundamental particles are produced with charge $\pm e$. Also, the particles are traveling at higher energies in which the energy losses are near or close to the minimum in the energy loss curve. The dynamic range of most TPC electronics can cover a wide range of particles in these types of experiments. In nuclear HICs, we are interested in measured particles with charges $Ze$ where Z=1-3, and even higher in some applications. The energy loss in the Bethe-Bloch equation, Eq.~\ref{eq:bb}, is proportional to $Z^2$. Even more, low energy nuclear HIC of intermediate beam energies (around \SI{300}{\MeVA}) produce low velocity particles which exist in the $1/\beta^2$ region of Eq.~\ref{eq:bb}, where the energy loss grows dramatically. The dynamic range of electronics significantly limit the PID as the charge of a particle increases and the velocity decreases where the electronics will be saturated.

Several TPCs have tried to address this issue by having regions of low and high gain, either in amplification gain or in electronics gain. This mitigates the complete loss of information but introduces a new problem. Though particles which deposit large amounts of charge will now have good measurements in the low gain areas, particles which deposit minimal energy losses will lose information in the same low gain areas. The reconstruction of such tracks will suffer. There are ongoing efforts in the nuclear community to develop new electronics which hope to mitigate these issues by developing more sophisticated  pre-amplifiers and electronics CITE HERE.  Never the less it is quite useful to develop a software technique which may extend the dynamic range of TPC electronics without the use of new external hardware, especially for experiments which have already been performed with older electronics technologies. In this section we will outline a novel software technique that takes advantage of the PRF described in Section~\ref{sec:prf} and can effectively extend the dynamic range of the TPC electronics.

In TPCs, the effective dynamic range is very different from the single channel dynamic range depending on how the TPC measurement is used. Typically TPCs are operated inside of a magnetic field for the purpose of reconstructing the momentum of a track, which requires sub-millimeter precision in the position determination of the track path. This is achieved by clustering several pads together as discussed in Section~\ref{sec:helixtrack}. To achieve this at least 2 adjacent pads must be measured, and the precision increases as the number of adjacent pads increases.

The effective dynamic range is then not the single channel dynamic range, but the dynamic range between central pad --holding the largest charge-- and the adjacent pads --holding the smaller charges in the PRF distribution. For example, to measure minimum ionizing particles, the Signal-to-Noise-Ratio (SNR) of the pad with the smallest charge in the distribution should be some reasonable value above noise, so that the signals can be measured. From the PRF in Section~\ref{sec:prf}, we know the central pad in the cluster holds 80\% of the total charge, where as the two adjacent pads each hold the remaining 10\%. In the \spirit TPC the electronics gain was set so that the the pads have a SNR of 6:1 for minimum ionizing tracks, and therefore the central pad had a SNR of 50:1. In the case of the maximum charge is collected in the central pad, before the electronics saturate, the SNR is 800:1. Therefore the maximum SNR is roughly 16 times that  --800 over 50 -- of minimum ionizing particles. This is order to have a complete measurement, where adjacent pads are all measured and position resolution is not compromised.

>>>>

The S$\pi$RITROOT software is modular code based on the FAIRROOT package, and is written in c++ \cite{fairroot}. The code is separated into modules which can be turned on or off depending on the users particular analysis. The base classes and frame work is provided by the FAIRROOT package, while the specifics of each detector need to be added. Here we will breifly describe the outline of the main tasks in the S$\pi$RITROOT software and then will discuss of each task later. The level of discussion concerning the software is really a practical, base understanding, appropriate for the discussion of this dissertation. For more details about the tracking algorithm the reader is refered to \cite{spiritroot_paper}, and to actual software which is published on GitHub \cite{spiritroot_git}.

@@

\subsection{Decoder task}
The decoder task reads in the raw data output from the GET electronics converting the binary data into \spirit ROOT classes using the GETDecoder \cite{getdecoder}. When the software is loaded, a map is loaded which maps each channel in the CoBo board onto the physical pad location on the pad plane. The information is stored in a pad container called the \emph{STPad} class. This container stores the row, layer, and the raw ADC time bucket spectrum; where the row refers to the pad numbering along the x-direction and the layer refers to the numbering along the z-direction. All of the pads combined form the event container called the \emph{STRawEvent} class which represents the raw event information.

Three basic calibrations are performed on the raw data level within this task. The gain calibration calibrates the electronics will be described in Section\ref{sec:elecCalib}. The anode gain calibration which will be discussed in Section~\ref{sec:anodeCalib}, and the gating grid noise subtraction and time bucket cut which is described in Section~\ref{sec:ggnoisesub}.
%gain calibration 
%ggnoise subtraction
%GG time bucket cut

\subsection{Pulse Shape Algorithm (PSA) Task}
\label{sec:psa}
%Yoffset match the TPC-Vertex_y with BDC_Y
%YPedestalOffset what is it for?
%cobo gain matching
An algorithm loops over the raw time bucket spectrum to find all the unique pules in each pad. It is very likely that there are several pulses in a pad coming from multiple tracks passing under the same pad which are separated only in arrival time. Figure~\ref{fig:psaTask} shows the PSA task working on one particular pad in experimental data. The shaded histogram in every sub-figure represents the ADC raw spectrum of the pad. In Fig.~\ref{fig:psa0} we see the PSA has identified the first pulse and fitted it with the standard pulse; the result of the fit is shown by the red line. The fitted red line is then subtracted from the ADC spectrum, which is shown in the shaded histogram in Fig.~\ref{fig:psa1}. Here, the second pulse is found and fitted again with the same procedure; the red line will be subtracted from the raw ADC spectrum. The black line represents the sum of all the fitted pules in each step. This continues until all pules are found or until the PSA reaches the last bin in the spectrum. Figure~\ref{fig:psa4} shows the final result where 4 fitted pulses were found. The agreement between the total summed line and the raw ADC spectrum exemplifies how successful the PSA is in this particular set of data.

@@

The fit of each individual pulse contains the height of the pulse and the timing information of the start of the pulse. We define the start of the pulse to be the location in time of 10\% of the rising edge. The height of the pulse is related to the charge induced on the pad, $Q$. Combined with the start time of the pulse $t_o$  it is saved into a new container representing a \emph{hit}, called the \emph{STHit} class. After the PSA processes each pad in the raw event, an entire hit cloud of the event is produced. The x and z-coordinates come from the center of the pad which is already stored in the map of STPad. Since the STHit class inherits from the STPad class this information is also carried over. The y-coordinate is calculated using the known drift velocity $v$ and time of the signal $t_o$ as $y = v\cdot t_0$, $t_0$. The hit class forms the fundamental structure from which all tracking is performed since it contains the x,y,z spacial coordinates, and charge information $Q$.

@@

 The function of the  helix track finding algorithm is to sort hit cloud into unique sets of hits representing tracks. This is performed by a Riemann track finding algorithm where the Cartesian space is mapped onto a Riemann sphere. This is a particular transformation in which unique helices in the Cartesian space form great circles on the Riemann sphere. This is particularly useful since hits which correspond to a track form a helices the Cartesian space of the TPC coordinates. It is difficult to search for tracks in the hit cloud by fitting a general helix to different combinations of hits. It is much easier to search for collection of hits by fitting a plane to the great circles which are formed on the Riemann sphere. More details on the Riemann tracking can be found here \cite{spiritroot_paper}. From here the set of hits which form a unique track are stored into a new helix track container called the STHelixTrack class.

%\paragraph{Definition of clustering}
The hits within a helix track are dense and the position resolution of an individual hit is not very precise. As mentioned in Section~\ref{sec:prf}, the TPC can achieve sub-millimeter position resolution by utilizing the PRF. In the helix track finding task, the hits are further reduced into more precise clusters. A brief description of the method of clustering is illustrated in Figure \ref{fig:topview}. As will be seen, it is sufficient to simply cluster hits along one direction, either along the x or z-direction, which ever is most perpendicular to the local track region. For example, the three clusters at the bottom of Figure \ref{fig:topview} are clustered along the x-axis, and the upper three are along the z-axis. The bold pads highlight the hits belonging to an example cluster from each set. Since we decide to cluster only in one direction, there is a natural inflection point in a track where the direction of clustering must switch. This depends on the crossing angle of the track $\theta$, which is defined as the angle between the tangent of the track -- at that cluster -- and the x-axis. In this definition, $\theta = \ang{90}$ corresponds to the case when the track is going along the z-axis and $\theta = \ang{0}$ for a track going along the x-axis.  The clustering direction in the case $\ang{45} < \theta \leq \ang{90} $ is along the x-direction. For tracks with angles $\ang{0} < \theta \leq \ang{45}$ we cluster along the z-axis.

@@

 The position along the clustering direction, $\overline{X}$, is calculated as the charge-weighted average, $\overline{X} = \sum_i q_ix_i$, where $q_i$ and $x_i$ are the charge and position of the i-th hit in the cluster. The other direction is set to the center of the pad. For example, if we are clustering hits along the x-axis, the z-position is set to the center of the pad in the z-direction and vice versa. Switching the clustering direction gives better position resolution than if we clustered only along one direction. You can imagine if we calculated the clusters only along the x-axis, tracks with $\theta \approx 0^{\circ}$ the x-position is not well defined. The cluster position and charge is stored into a new container representing the cluster called the STHitCluster class. All the clusters which belong to a particular helix track is also stored in the corresponding STHelixTrack class.

\subsection{Correction Task}
%extend dynamic range
%PRF cut
%space charge correciton
In this task the hit distribution, within a given cluster, is checked against the measured PRF discussed in Section~\ref{sec:prf}. Figure~\ref{fig:prfpimData} shows black lines around the measured PRF in the data representing gates defining the acceptable PRF region. If less than 50\% of the hits within a cluster lie within this gate, the cluster is thrown out from the track. A majority of the clusters contains either 2 or 3 hits in total.  Clusters which do not follow this PRF gate usually have been corrupted in some way. Typically this occurs when charges from other nearby tracks mix distorting the hit distribution which corrupts the energy loss and position of the cluster. This tends to occur most frequently in the dense region surrounding the target, where the track density is high. But it may occur along the track for any other reason. The width of the gate is to allow for a reasonable range of PRF values to be accepted, without biasing the data too much.

@@

This correction task also handles several other important corrections. The first is a novel algorithm which extends the dynamic range of the electronics and is discussed in detail in Section~\ref{sec:extendDynamicRange}. The other correction task that is handled here is the space charge correction which will be discussed in more detail in Section~\ref{sec:spacecharge}. These two corrections represent the most significant corrections that greatly improve the data.

\subsection{Momentum and track reconstruction (GENFIT)}
After all of the corrections are applied to the clusters, the cluster's position and charge values are then passed to a software package which reconstructs the momentum of the track called GENFIT. GENFIT utilizes a Kalman fitting algorithm which returns the estimated momentum and charge values for a given track \cite{genfit}. These values are then stored in into a new track container called the STRecoTrack class.

This task is repeated a second time, but this time including the vertex as a constraining point in which all tracks are refitted. In this way, the momentum resolution is greatly improved by adding the vertex as a constraining point; since the BDC detectors, described in Section~\ref{sec:bdc}, give such an accurate vertex location. The energy loss of the track is calculated by the truncated mean method described in Eq.~\ref{eq:truncmean}. The GENFIT task represents the final part of the analysis which pertaining to track reconstruction. From here the PID can be constructed using the energy loss, momentum, and charge information.

\subsection{Vertex tracking (RAVE)}
\label{sec:vertex}
After all tracks are reconstructed by GENFIT, the tracks are then passed to the RAVE software package which reconstructs the event vertex from a collection of tracks \cite{rave}.  The RAVE vertex is typically referred to as the TPC vertex in this dissertation. It is used primarily to find if an event is on target by observing the z-distribution as will be discussed in Section~\ref{sec:vertexcut}. The vertex which is used in refitting the tracks as mentioned in the above section comes from the BDC interpolated vertex which is more accurate than the TPC generated vertex.

@@

\subsection{Gating grid noise subtraction}
\label{sec:ggnoisesub}
Opening the gating grid is essentially done by shorting the adjacent wires in the grid together, and allowing them to reach equilibrium as fast as possible. Since the impedance of both sides was not entirely matched, this caused an imbalance of charge where one side discharged faster than the other side, causing charge to bounce back and forth in the grid in an under-dampened system. This create oscillating residual noise early in the time bucket spectrum. Figure~\ref{fig:ggNoiseSub} shows in the upper panel shows the ADC time bucket spectrum of the gating grid noise for 2000 events in a given pad. The gating grid noise is stable for a given pad and the mean value --shown as the red line-- can be calculated after averaging over several thousands events. 
 %The ground grid shielded some of the noise, but the path to ground was on two ends and was not sufficiency for this frequency.

@@

The bottom panel of Fig.~\ref{fig:ggNoiseSub} shows the gating grid noise self correcting itself, representing a best case scenario. The pedestal -- or base line ADC-- is around 3800 ADC in the upper panel, is automatically subtracted when subtracting the gating grid noise profile since the subtracted ADC spectra is centered around 0. Still some residual gating grid noise remains even after the subtraction, but it is significantly reduced and its extent is over a much smaller time bucket region. This failure happens in regions where the derivative of the gating grid noise is large. In these areas the ADC value is dependent upon small jitters in the timing of the signal and as a result varies wildly. To take care of the residual noise in the subtracted spectra we remove all bins below 30 Tb.

In the experiment the gating grid itself was quite stable in the  ${}^{132}$Sn and ${}^{124}$Sn systems and was not stable in the ${}^{108}$Sn and ${}^{112}$Sn beams, where the it broke many times. Because of this, the gating grid noise was monitored regularly and we took gating grid noise profiles frequently, especially after a gating grid board was replaced.

@@

A cocktail of several light charged particles --at a known energy-- was produced (p,d,t,${}^{3}$He,${}^{4}$He) and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BigRIPS spectrometer were set so that the momentum resolution of the beam was known to within $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied with no target in place. A thick Aluminum target was also inserted to provide a slightly lower energy calibration point for part of the lower rigidity setting, effectively creating three energy calibration points over several particle species. In some cases the certain particles were not able to propagate or were strongly attenuated and provided poor statistical measurements.

Since the cocktail beam was a mono-energetic source of particles and the actual momentum resolution was less than 1\%, the observed momentum resolution can be interpreted as the intrinsic TPC momentum resolution. The momentum resolution in general depends on several factors such as the particle's angle, momentum, charge, track multiplicity, etc. This calibration beam represents an ideal situation where the track was parallel to the pad plane and the track is not affected by the presence of other tracks. The average momentum resolution was measured to be $dp/p = 2\%$ over the full range of particle species for all settings as measured by the sigma of the Gaussian momentum distribution.

The energy loss resolution can also be directly measured a monochromatic source of each particle species have a well defined energy loss distribution. The energy loss resolution was measured to be 5\% for all particles over the energy setting measured and is summarized in Table~\ref{tb:momresolution}.

Since the magnetic dipole setting of the BigRIPS spectrometer is well defined, we can calculate the expected momenta of each particle species measured. The enegy losses through various materials in the beam line were propagated using LISE++ software \cite{lise++}. The measured momenta of the calibration beam differed significantly from the expected values -- up to 5\% in the high momentum calibration -- as seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. This effect is attributed to inhomogenatities in the magnetic field which introduces electron drift velocities in the direction of $\vec{E}\times\vec{B}$ direction as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality.  The details of the correcting for the $\vec{E}\times\vec{B}$ effect will be discussed later in Section~\ref{sec:spacecharge} in a more general way which also includes the correction for the presence of positive space charge in the field cage volume. The same correction technique was applied here, where the cocktail beam is the special case of zero space charge.

The values under the corrected column in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}, represent the data correcting for the $\vec{E}\times\vec{B}$ effect. A significant improvement is seen in the high momentum setting going from around ~5\% disagreement to within ~1\% agreement in the corrected data. For the lower momentum settings (Tables~\cref{tb:cocktail100tar,tb:cocktail100}), protons see a slight improvement of about ~1\% where as the deuterons are over corrected in both settings. The level of agreement of the all corrected values is still well within the estimated momentum resolution of the TPC.

@@

The electronics were calibrated by measuring the response of each channel to an input signal supplied by a pulse generator. This is a relative calibration technique with the intent to calibrate the varying gains in each channel relative to one another, not an absolute calibration. The pulse was distributed to all the electronics channels by pulsing the ground plane for a range of input voltages. This distributed the pulse evenly across the entire pad plane. The input voltage is plotted as a function of the measured height of the pulse in the electronics channel given in units of ADC in Fig.~\ref{fig:gaincalib}, for every channel. The small variation in each channel can be seen as the wide band around each measurement point. A linear fit is performed to get the calibration line for each channel. A particular channel is chosen which provides a reference calibration to which the slopes of all other channels are calibrated too.   The right panel shows the resulting distribution of channels after calibration, in which the channel variation is reduced significantly.

\subsection{Anode gain calibration}
\label{sec:anodeCalib}
As discussed in Section~\ref{sec:wireplanes} the voltage of the anode sections 12 and 14 were reduced during the experiment due to high currents being observed on the wires. Out of all the runs used in the analysis in this thesis -- see \ref{tb:runList} -- the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, we artificially increase the gain of the pads which lie above these anode wires in the software. The multiplicative factor is estimated by plotting the energy loss in the high gain sections relative to the low gain sections.

@@

 A map defining the gain calibration of each pad is pre-loaded into the software. In the decoder task the raw ADC spectra was multiplied by this gain factor for each channel. The PSA thresholds of those channels also were multiplied by the same gain factor since the noise levels were also artificially magnified as well.

@@

Using a TPC for measurements of HICs in nuclear physics presents a different set of challenges as opposed to higher energy experiments. Typically in higher energy experiments fundamental particles are produced with charge $\pm e$. Also, the particles are traveling at higher energies in which the energy losses are near or close to the minimum ionization in the energy loss. In these cases the dynamic range of most TPC electronics can cover a wide range of particles. In nuclear HICs, we are interested in measured particles with charges $Ze$ where Z=1-3, and even higher in some applications. Recall the energy loss in the Bethe-Bloch equation, Eq.~\ref{eq:bb}, is proportional to $Z^2$. Even more, in low energy nuclear HIC of intermediate beam energies -- around \SI{300}{\MeVA} -- produce low velocity particles which exist in the $1/\beta^2$ region of Eq.~\ref{eq:bb}, where the energy loss grows dramatically. In this case, the limited dynamic range of electronics significantly limit the PID as the charge of a particle increases and the velocity decreases, where they saturate the electronics.

Several TPCs have tried to address this issue by having regions of low and high gain, either in amplification gain or in electronics gain. This mitigates the complete loss of information but introduces a new problem. Particles which deposit large amounts of charge will have good measurements in the low gain areas, where as particles depositing minimal energy losses will lose information in the same low gain areas. The reconstruction of such tracks will suffer. There are ongoing efforts in the nuclear community to develop new electronics which hope to mitigate these issues by developing more sophisticated  pre-amplifiers and electronics CITE HERE.  Never the less, it is useful to develop a software technique which extends the dynamic range of TPC electronics without the use of new external hardware. Which can be particularly useful for experiments which have already taken data with older electronics technologies. In this section we will outline a novel software technique that takes advantage of the PRF described in Section~\ref{sec:prf} and can effectively extend the dynamic range of the TPC electronics.

In TPCs, the effective dynamic range is very different from the single channel dynamic range depending on how the TPC measurement is used. Typically TPCs are operated inside of a magnetic field for the purpose of reconstructing the momentum of a track, which requires sub-millimeter precision in the position determination along the track path. This is achieved by clustering several pads together as discussed in Section~\ref{sec:helixtrack}. To achieve this at least 2 adjacent pads must be measured, and the precision increases as the number of adjacent pads increases.

In this case the effective dynamic range is not the single channel dynamic range, but the relative dynamic range between central pad --holding the largest charge-- and the adjacent pads --holding the smaller charges in the PRF distribution. For example, to measure minimum ionizing particles, the Signal-to-Noise-Ratio (SNR) of the pad with the smallest charge in the distribution should be some reasonable value above noise, so that the full charge distribution can be measured. From the PRF in Section~\ref{sec:prf}, we know the central pad in the cluster holds about 80\% of the total charge, where as the two adjacent pads each hold the remaining 10\%. In the \spirit TPC the electronics gain was set so that the the pads have a SNR of 6:1 for minimum ionizing tracks, and therefore the central pad had a SNR of 50:1. The maximum SNR in the central pad -- before the electronics saturate -- the SNR is 800:1. Therefore the maximum SNR is roughly 16 times that  --800 over 50 -- of minimum ionizing particles.
